Empirical Bayes analysis for detection of gene heterosis in RNAseq data

### 2014-12-17 ebayes_rnaseq_heterosis.tex
I rewrote the paragraphs on edgeR and baySeq.

% WML: change: "edgeR was originally designed to..."

{\tt edgeR} was originally designed to test for differential expression between two varieties \citep{robinson2007moderated, robinson2010edgeR}, and we used this capability to construct a statistic for each gene. If the hybrid's mean expression does not differ from a parent, then the hybrid does not exhibit heterosis. We can calculate two pvalues: one for the null hypothesis that the hybrid's mean expression is the same as parent 1 and the second for the null hypothesis that the hybrid's mean expression is the same as parent 2. But, even if both pvalues are small, the hybrid's mean expression could be between the two parents, i.e. mid-parent heterosis, and therefore not be exhibiting extreme heterosis. We construct a statistic that is 0 if the estimate of mean expression 
for the hybrid is between the estimates for the parents. Otherwise, the statistic is one minus the pvalue above corresponding to whichever parent's mean estimate is closer to that of the hybrid.

@@

The {\tt baySeq} approach allows for a wider range of hypotheses including $H_1: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H_2: \mu_{g1}=\mu_{g2}$, $H_3: \mu_{g1}=\mu_{g3}$, $H_4: \mu_{g2}=\mu_{g3}$, and $H_5:$ all $\mu_g$'s are distinct \citep{hardcastle2010bayseq, hardcastle2012baySeq}. Hypotheses $H_2$ and $H_5$ are the only two that indicate the possibility of heterosis. Similar to the {\tt edgeR} approach, we construct a statistic that is 0 if the estimated hybrid mean, based on the MLE's from {\tt edgeR}, is between the parental means and otherwise is the sum of the posterior probabilities for hypotheses $H_2$ and $H_5$.

>>>>

% WML: I rewrote the paragraph on edgeR.

{\tt edgeR} is designed to apply a negative binomial loglinear model to test for differential expression between any two varieties \citep{robinson2007moderated, robinson2010edgeR}, and we used this capability to construct a pseudo-probability of heterosis for each gene. We first computed the maximum likelihood estimates of the mean expression effects of all genes using {\tt edgeR}'s built-in Fisher scoring algorithm, and then we used the likelihood ratio test functionality to calculate two legitimate p-values for each gene: one for the null hypothesis that the hybrid's mean expression is the same as that of parent 1 and the second for the analogous test involving parent 2. For each gene, if the estimated mean expression level of the hybrid is not between the estimated mean expression levels of the parents, we set the pseudo-probability of heterosis to be one minus the p-value corresponding to the parent with estimated expression level closer to that of the hybrid. Otherwise, the pseudo-probability is zero.

@@

% WML: I rewrote the baySeq paragraph.

The {\tt baySeq} approach allows for a wider range of hypotheses for each gene, including $H_1: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H_2: \mu_{g1}=\mu_{g2}$, $H_3: \mu_{g1}=\mu_{g3}$, $H_4: \mu_{g2}=\mu_{g3}$, and $H_5:$ all $\mu_g$'s are distinct \citep{hardcastle2010bayseq, hardcastle2012baySeq}. In a technique similar to our application of {\tt edgeR}, we used used the posterior probabilities of these hypotheses to construct a pseudo-probability of heterosis for each gene. We set this pseudo-probability to zero if the hybrid's maximum likelihood estimate of the mean expression level calculated with {\tt edgeR} is between those of the parents. Otherwise, the pseudo-probability is the sum of the posterior probabilities of $H_2$ and $H_5$. Note that $H_2$ and $H_5$ are the only hypotheses that allow for heterosis.
